Abstractions for Type-Level Programming

### 2022-04-22 article.tex
Make it clear
Talk about co- and contravariance of MT

\subsection{Variance}

@@

Scala tuples, for example, fall into this category.
|Tuple2|, the class for pairs, is defined as follows:
%
\begin{lstlisting}
case class Tuple2[+T1, +T2](_1: T1, _2: T2)
\end{lstlisting}
%
Given two instantiations of this this type, |Tuple2[Int,X]| and |Tuple2[String,X]|, although |Tuple2[Nothing,X]| is a subtype of both, there is no runtime value of type |Tuple2[Nothing,X]| (since |Nothing| is uninhabited), and, as a result, those two types are disjoint.
Our disjointness algorithm implements this kind of reasoning to conclude disjointness in the presence of covariant type parameters.

% TODO make it clear that contravariant types are never considered disjoint.
% TODO Incorporate email discussion on match types co- and contravariance.

>>>>

\subsection{Disjointness of Variant Types}
\label{subsec:disjointness-of-variant-types}

@@

Scala tuples, for example, fall into this category.
|Tuple2|, the class for pairs, is defined as follows:
%
\begin{lstlisting}
case class Tuple2[+T1, +T2](_1: T1, _2: T2)
\end{lstlisting}
%
Given two instantiations of this this type, |Tuple2[Int,X]| and |Tuple2[String,X]|, although |Tuple2[Nothing,X]| is a subtype of both, there is no runtime value of type |Tuple2[Nothing,X]| (since |Nothing| is uninhabited), and, as a result, those two types are disjoint.
Our disjointness algorithm implements this kind of reasoning to conclude disjointness in the presence of covariant type parameters.
There is no such reasoning available for contravariant types: our algorithm can never conclude that |G[X]| and |G[Y]| are disjoint for a contravariant type |G|.

\subsection{Match Type Variance}

In \Cref{subsec:disjointness-of-variant-types} we discussed how our match type reduction algorithm handles variance annotation of higher-kinded types.
In this section, we discuss the orthogonal problem of supporting variance annotations on match type definitions.

At the time of writing, our implementation does not support variance annotations on match types (all match types are invariant).
However, it would be conceivably to lift this restriction in the future, as shown by the analysis presented in this section.

Consider the following definition of a match type |M|:
%
\begin{lstlisting}
type M[X] = X match
  case P1 => E1
  case P2 => E2
  case P3 => E3
\end{lstlisting}
%
In \Cref{sec:formalization}, we define the semantics of this type in terms of subtyping rules.
Given this particular definition of |M|, we can write a direct semantic definition of |M| with equations:
%
\begin{itemize}
 \item |M[X] = E1| \quad if |X <: P1|
 \item |M[X] = E2| \quad if |disj(X, P1)| and |X <: P2|
 \item |M[X] = E3| \quad if |disj(X, P1)| and |disj(X, P2)| and |X <: P3|
\end{itemize}
%
This equational definition is consistent under the assumptions of \Cref{lem:disjointness-subtyping-exclusivity}.

We show that |M[X]| is covariant if |E1|, |E2|, and |E3| are covariant with respect to |X|.
Given |T1 <: T2|, we proceed by case analysis on |M[T2]|:
%
\begin{itemize}
  \item If |M[T2] = E1|, then |M[T1] = E1| by subtyping transitivity and the result follows from the covariance of |E1|.

  \item If |M[T2] = E2|, then (|disj(T2, P1)| and |T1 <: T2|) implies |disj(T1, P1)| (using D-SUB), and (|T2 <: P2| and |T1 <: T2|) implies |T1 <: P2|, which yields |M[T1] = E2|.
  The result follows from the covariance of |E2|.

  \item |M[T2] = E3| is analogous.
\end{itemize}
%
Similarly, we can show that |M[X]| is contravariant if |E1|, |E2|, and |E3| are contravariant with respect to |X|.

In practice, it would be hard to show that |E1|, |E2|, and |E3| are co-/contravariant when type bindings get into the picture. For non-binding cases, that analysis is straightforward and analogous to the variance checks that are already in place in the compiler.
